\relike: Reionization Effective Likelihood from Planck 2018 Data

### 2021-04-28 paper/rei.tex
Final copy

@@

\title{RELIKE: Reionization Effective Likelihood from Planck 2018 Data}

@@

We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC). This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested. We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation. We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler. We expect the code release to dramatically reduce the time for testing individual global ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as \ch{line-intensity mapping},
%21cm,
luminosity function, star formation history, etc.
Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.). The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}

@@

\ch{CH: going through the paper once from beginning to end now.}

\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}. mean free path of ionizing photons  between 5 and 6: \url{https://arxiv.org/pdf/2103.16610.pdf}, $<80\%$ ionized at z = 6. }

The cosmic microwave background (CMB) has entered an era of precision cosmology. The measurements from the Planck satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution~\cite{Aghanim:2018eyx}. While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein). Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.

Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced. This \ch{steplike tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition. However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.

In fact, to extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, for studying inflationary features vs reionization features in Ref.~\cite{Obied:2018qdr}, as well as massive neutrinos and gravitational waves in Ref.~\cite{Dai:2015dwa}. %\cite{Obied:2017tpd}

In a re-analysis of the Planck 2015 data with PCs~\cite{Heinrich:2016ojb}, a component of the high-redshift ionization that would have been missed by a simple steplike model~\cite{Heinrich:2016ojb} was uncovered. In the latest official Planck 2018 release~\cite{Aghanim:2018eyx}, the PC method was also adopted to probe ionization at high-redshift, whose significance was reduced since the Planck 2015 release, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Heinrich:2018btc}. In addition to PCs, the FlexKnot method was employed, which is also able to capture general ionization histories by varying the number of ``knots" in redshifts and the amplitude of the ionization fraction at these knots. 
% Planck intermediate results on reionization history: \cite{Adam:2016hgk}

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
~\cite{Delouis:2019bub}, adopts better foreground modeling techniques and allowing for improved reionization constraints.
%
%The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.
%\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
%
\ch{Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data. In this paper, we obtain new reionization PC constraints with this latest likelihood $\texttt{SRoll2}$. Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood code useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}. 
}

\ch{We publish this reionization effective likelihood code (RELIKE) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}. There are two packages that come with the code release: 1) we publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model. If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space. 2) We also provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in the package \texttt{CosmoMC-relike}. We expect that this code release would allow much simpler and faster testing of specific models in the Planck data, and to provide consistent and fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity functions, star formation history, 21cm and line-intensity mapping in general, etc.
}

Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release. But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is about \ch{$\sim 3$ times} less stringent than the Planck official results using Flex Knots in Ref.~\cite{Aghanim:2018eyx}. Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
~\cite{Millea:2018bko} and ends up penalizing lower optical depth models. [quote difference in tau(15, 30) or total tau possibly attributed to this treatment]. %We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data

The paper is structured as follows. We first describe in \S\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood. Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{SRoll2} likelihood, and compare with previous results in literature. Then in \S\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes. Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. Here, we  briefly summarize the PC technique itself in \S\ref{sec:PC}.   We present  techniques 
%kernel density estimate technique
for building its effective likelihood which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
%Hu:2003gh, %Mortonson:2008rx, 
Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)} = \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected 
%
% added ``expected" to distinguish this from sigma(ma)
%
variances of the PCs.
The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.

We then rank-order the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by  taking 
 for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}

 In Fig.~\ref{fig:xe}
 %
 % this seems to be in the main text now
 %\footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
 %
 (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
 \beq
 f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
 \eeq
as the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.

We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using COSMOMC\footnote{COSMOMC: \url{http://cosmologist.info/cosmomc}} of the relevant Planck likelihood discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density 
$\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
its tilt $n_s$.  We take flat uninformative
priors in $m_a$ but discuss additional priors imposed by
removing demonstrably  unphysical ionization histories in \S \ref{sec:cumulative}.

 \begin{figure}
          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume fully ionized Universe at $z < 6$. \textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC. } 
          \label{fig:xe}
\end{figure}

@@

An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{n_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%

For any set of model parameters ${\bf p}$ with their given prior probability distribution
$P({\bf p})$, their posterior distribution can be approximated as 
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
\end{equation}
with PC\,=\,KDE or Gaussian.
This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
In \S\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

%
%
%
%We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.

\subsection{Cumulative Optical Depth}
\label{sec:cumulative}

>>>>

\newcommand{\relike}{RELIKE}%\texttt{RELIKE} 
\newcommand{\cosmomcrelike}{\texttt{CosmoMC-relike}}

@@

\title{\relike: Reionization Effective Likelihood from Planck 2018 Data}

@@

We release \relike\  (Reionization Effective Likelihood), a fast and accurate effective likelihood code based on the latest Planck 2018 data 
that allows one constrain any model for reionization between $6 < z < 30$ using five constraints from the CMB reionization principal components (PC). We tested the code on two example models which showed excellent agreement with sampling the exact Planck likelihoods using either a simple Gaussian PC likelihood or its full kernel density estimate.
This code enables a fast and consistent means for 
combining Planck constraints with other reionization data sets, such as kinetic Sunyaev-Zeldovich effects, line-intensity mapping, luminosity function, star formation history, quasar spectra, etc, where the redshift dependence of the ionization history is important.  
Since the PC technique tests any reionization history in the given range, we also derive model-independent constraints for the total Thomson optical depth $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and its $15\le z \le 30$ high redshift component $\tau_{\rm PC}(15, 30) < 0.020 $ (95\% C.L.).
The upper limits on the high-redshift optical depth is a factor of $\sim3$ larger than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method
and we validate our results with a direct analysis of a two-step model which permits this small high-$z$ component.

@@

The cosmic microwave background (CMB) has helped establish the $\Lambda$CDM model as the standard cosmological model~\cite{Aghanim:2018eyx}. While many components of this standard cosmology model are well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein).  It is usually characterized by a single parameter, the total Thomson optical depth, whose  uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude. Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.

Many probes of  the reionization era exist today and with future probes it will only become more important to combine them consistently the full implications of the CMB. In the CMB data, free electrons during the reionization epoch Compton scatter with the CMB photons to cause a suppression of the CMB primary anisotropies; on the large scales, the same scattering process of the CMB temperature quadrupole in the frame of the free electrons induces an additional reionization bump in the CMB $E$-mode polarization. On the small scales, the kinetic Sunyaev-Zeldovich (kSZ) effects generated by the scattering of the CMB photons off ionized gas in bulk motion~\cite{zeldovich_sunyaev_1969, sunyaev_zeldovich_1980} are used to constrain morphology and duration of reionization~\cite{mcquinn_2005,mesinger_2012_kSZ}. Additionally, observations of galaxy luminosity functions, quasars, star formation rate, line-intensity mapping also provide important and complementary information on reionization~\cite{2016ASSL..423.....M}. In this paper, we aim to extract all the information in the large-scale $E$-mode observations for constraining the global ionization history of the Universe, and provide a way to consistently analyze them with all of these other probes.

In the standard CMB analysis, reionization has been  modeled as a steplike transition in the global ionization history in the form of a tanh function, with the step location parameterized by the total Thomson optical depth induced. This steplike tanh model assumes, by construction, that there is negligible ionization before the transition. However, the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles contains more information on the coarse-grained evolution of the ionization history than the total optical depth~\cite{Hu:2003gh, Mortonson:2007hq}.

To extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum. This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}. It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, for studying inflationary features vs reionization features in Ref.~\cite{Obied:2018qdr}, as well as massive neutrinos and gravitational waves in Ref.~\cite{Dai:2015dwa}. %\cite{Obied:2017tpd} 
In a re-analysis of the Planck 2015 data with PCs~\cite{Heinrich:2016ojb}, a component of the high-redshift ionization that was missed by a simple steplike model was mildly preferred.

In the latest official Planck 2018 release~\cite{Aghanim:2018eyx},  this mild preference for finite high redshift ionization in the Planck 2015 release turned into upper limits, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Millea:2018bko} resulting in reduced $C_l^{EE}$ around $l \lesssim 10$ (see also Ref.~\cite{Heinrich:2018btc}). The FlexKnot method was primarily employed, which is also able to capture general ionization histories by varying the number of ``knots" in redshifts and the amplitude of the ionization fraction at these knots.

Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}, adopts better foreground modeling techniques and allowing for improved reionization constraints \cite{Delouis:2019bub}.

Because the Planck data sets the current standard for CMB constraints, it is important to extract all the information present on reionization  so that it may be readily used in future joint analyses with other cosmological probes. In this paper, we obtain new reionization PC constraints using legacy Planck 2018 likelihood and this latest likelihood $\texttt{SRoll2}$ for the low-$\ell$ $E$-mode power spectrum.

Enabled by the completeness property of the PCs, we turn these constraints into an effective means of  assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30. 
The 95\% C.L. upper limit on the cumulative optical depth between $z=15$ and 30 is $\tau(15, 30) < 0.020$  and about $\sim3$ times less stringent than reported using FlexKnot in the official Planck analysis $\tau(15, 30)<0.007$~\cite{Aghanim:2018eyx}.  We validate our results with an explicit toy model that allows such a small component of high redshift ionization and show that this result is not simply a consequence of principal component priors alone as suggested by \cite{Millea:2018bko} nor should these priors be changed as was adopted in the official Planck analysis~\cite{Aghanim:2018eyx}.
Extending the range of PCs to cover up to $z_{\rm max} = 50$, we verify that our results are robust to this extension.

With the technique  validated,
we release the Reionization Effective LIKElihood code (\relike) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}. There are two packages that come with this code release: 1)  a python package \relike, to quickly evaluate the effective Planck likelihood of any reionization model given $x_e(z)$. If desired, this package can be connected to a MCMC sampler to sample the posteriors in a higher-dimensional parameter space. 2)  the CosmoMC sampler with a fortran implementation of \relike\  code in the package \cosmomcrelike. This code enables a  much simpler and faster testing of specific reionization models with the Planck data, and provides a  consistent way of combining this information with other reionization datasets.

The paper is structured as follows. We first review in \S\ref{sec:background} the methodology of  reionization principal components along with the kernel density estimate (KDE) technique and the Gaussian approximation \cite{Heinrich:2016ojb} used for building the effective likelihoods. Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{SRoll2} likelihoods, and compare with previous results in literature. In \S\ref{sec:effective_likelihood}, we test the effective likelihood code and demonstrate its fidelity with two examples: 1) the standard steplike model; and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard tanh model for illustration purposes. Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.

@@

\section{PC Technique}
\label{sec:background}
The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}. Here, we  briefly summarize the PC technique itself in \S\ref{sec:PC}. We present  techniques for building its effective likelihoods which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.

@@

\subsection{Reionization Principal Components}
\label{sec:PC}
We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
%
\begin{equation}
x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
\label{eq:mmutoxe}
\end{equation}
%
where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model. We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements 
%
\beq
F_{ij} = \sum_l \frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)} = \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
\eeq
%
where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected variances of the PCs.
The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed 
 with Planck 2015 rather than WMAP best-fit,
 which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.

We then rank-order the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit. In the work that follows, we therefore truncate the sum at $a = 1 .. N_{\rm{PC}}$, where $N_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by taking 
 for $z_{\rm max} = 50$ where $N_{\rm{PC}} = 7$ is required at the cosmic variance limit.}

 In Fig.~\ref{fig:xe} (top), we show the corresponding fiducial basis functions $S_a(z)$.  
Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
As a consequence  the $N_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
given any $x_e^{\rm true}(z)$, we can project it onto the $N_{\rm PC}$ PC basis through
\begin{equation}
m_{a}=
  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
\label{eq:xetommu}
\end{equation}
%
where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision. We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $N_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.   However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.

 We  compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh} (see \cite{Heinrich:2016ojb}).
 %\todo{CH: probably meant \cite{Heinrich:2016ojb}}.
 We follow CAMB and assume for $z<6$, fully ionized hydrogen and singly ionized helium whereas for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
 \beq
 f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
 \eeq
as the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.

We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using CosmoMC\footnote{CosmoMC: \url{http://cosmologist.info/cosmomc}} \todo{CH:cite} of the relevant Planck likelihoods discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density 
$\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
its tilt $n_s$.  We take flat uninformative
priors in $m_a$ but discuss additional priors imposed by
removing demonstrably  unphysical ionization histories in \S \ref{sec:cumulative}.

 \begin{figure}
          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$. We assume a fiducial model with fully ionized hydrogen and singly ionized helium by $z<6$, and doubly ionized helium at $z = 3.5$.
          \textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC. } 
          \label{fig:xe}
\end{figure}

@@

An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
%
\begin{equation}
 {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
 \label{eq:gaussian}
 \end{equation}
%

For any set of model parameters ${\bf p}$ with their given prior probability distribution
$P({\bf p})$, their posterior distribution can be approximated as 
\begin{equation}
P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
\end{equation}
with PC\,=\,KDE or Gaussian.
This approach obviates the need to implement specific reionization models into CAMB and conduct separate CosmoMC sampling for each, thereby significantly reducing both human and computational effort.
In \S\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.

%
%
%
%
\subsection{Cumulative Optical Depth}
\label{sec:cumulative}

Although reionization PCs are mainly a tool for model testing using complete information from large angle CMB polarization, as they do not reconstruct the rapidly varying aspects of $x_e(z)$ itself, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
\begin{align}
\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
\label{eq:cumtau}
\end{align}
where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
\begin{align}
\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^{N_{\rm PC}} m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
\label{eq:cumtauPC}
\end{align}
where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
Eq.~(\ref{eq:cumtau}) using the fiducial model 
for $H(z)$ and number densities.  This differs from Ref.~\cite{Heinrich:2016ojb} where their cosmological dependence was retained in determining $\tau_{\rm PC}$.
 When we omit the redshift arguments, we
implicitly mean the total range, e.g. 
 $
 \tau_{\rm PC}\equiv \tau_{\rm PC}(0,z_{\rm max}).
 $

 In Fig.~\ref{fig:xe} (bottom), we display $\tau_a(z,z_{\rm max})$ for the $\zmax=30$ PCs.
 Notice that positive values of the first component $m_1$ mainly represents the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.    Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
The underlying reason is that in $C_l^{EE}$ the higher the redshift, the
larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.  For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth and multipole for the underlying $C_l^{EE}$ spectrum.
